Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-20 paper/paper.tex
Revise the text about de bruijn related functions (WIP)

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{if} \\ \\
\mathit{Renaming}\\
[[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang \scw{Where does the name come from?}}
  \label{fig:syntax}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}.
\scw{Why so many metavariables for terms? Do you need $c$, $p$ and $t$?}
We use
the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper.  \scw{Give an overview of the figure}
\scw{Define a convention that separates your use of natural numbers. Sometimes they
are universe levels and sometimes they are de Bruijn indices. Better to have separate
metavariables for each. (And it is possible to generalize universe levels to structures other than natural numbers. Ask Jonathan.}
We find de Bruijn
representation advantageous for specification since it leaves very
little ambiguity about the variable freeness side
conditions, making our proof more easily reproducible. Furthermore, as we discuss in
Section~\ref{sec:automation}, de Bruijn representation is much more
amenable to automated reasoning in proof assistants.
We adopt from \citet{autosubst} the notations for simultaneous
renaming, substitution, and other auxiliary definitions related to our
syntax, summarized in Figure~\ref{fig:auxdef}.
\scw{Need to explain the notations in the text. What are these operations?
What do readers need to understand about them? Help me understand the figure}
\scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
indices work.}
We omit most of the
definitions of renaming and substitution and only show the definition
of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        \mathit{IdentityRen} \\
        [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm i]] & := & [[var i]] \\ \\
        \mathit{ConsRen ([[xi .: n]])} \\
        ([[xi .: n]])([[0]]) & := & [[n]] \\
        ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n i]] & := & [[n]] + [[i]]
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \mathit{UpRen} \\
        [[( up xi ) 0]] & := & [[0]] \\
        [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        \mathit{Renaming ([[a < xi >]])} \\
        [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        \ldots \\ \\
        \mathit{UpTm \scw{shifting? lifting?}} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var i { rho }  ]] & := &  [[rho i]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        && \fbox{ \scw{case for $\lambda$ } } \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

% Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
% syntax, including renaming, substitution, and operations over
% the typing context or substitution.
As a
dependent type theory, terms and types are collapsed into the same
syntactic category. Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable. $[[Set i]]$ represents the universe
type where $[[i]]$ ranges over the set of natural numbers.
 Finally,
\lang\scw{Don't begin a sentence with a symbol} also includes an intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.
\scw{Example of what you can use identity types for? Or an example of
a program that uses J?}

\lang is expressive enough to support large
eliminations, the ability to compute a type using a term as input. For
example, the function $[[\ Bool if var 0 Bool Bool -> Bool]]$ returns
either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
is $[[true]]$ or $[[false]]$.

>>>>

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{if} \\ \\
% \mathit{Renaming}\\
% [[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang \scw{Where does the name come from?}}
  \label{fig:syntax}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}.
\scw{Why so many metavariables for terms? Do you need $c$, $p$ and $t$?}
We use
the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper since it is more amenable to
automation (Section~\ref{sec:automation}) and leaves no ambiguity
about variable freeness conditions.

 \scw{Give an overview of the figure}
\scw{Define a convention that separates your use of natural numbers. Sometimes they
are universe levels and sometimes they are de Bruijn indices. Better to have separate
metavariables for each. (And it is possible to generalize universe levels to structures other than natural numbers. Ask Jonathan.}
As a dependent type theory, terms and types are collapsed into the same
syntactic category. The type $[[Set i]]$ represent universe
types and $[[var n]]$ represent de Bruijn variables. While
$[[i]],[[j]],$ and $[[n]]$ are all metavariables representing natural
numbers, we always use $[[n]]$ to represent term variables that
participate in substitution and $[[i]],[[j]]$ to represent universe levels.
Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable.
Finally, we include in \lang{} the intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.
\scw{Example of what you can use identity types for? Or an example of
a program that uses J?}

% We find de Bruijn
% representation advantageous for specification since it leaves very
% little ambiguity about the variable freeness side
% conditions, making our proof more easily reproducible. Furthermore, as we discuss in
% Section~\ref{sec:automation}, de Bruijn representation is much more
% amenable to automated reasoning in proof assistants.

\scw{Need to explain the notations in the text. What are these operations?
What do readers need to understand about them? Help me understand the figure}
\scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
indices work.}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm n]] & := & [[var n]] \\ \\
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n i]] & := & \ottkw{v}_{[[n]] + [[i]]}
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        % \mathit{UpRen} \\
        % [[( up xi ) 0]] & := & [[0]] \\
        % [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        % \mathit{Renaming ([[a < xi >]])} \\
        % [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        % [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        % \ldots \\ \\
        \mathit{Lifting} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var i { rho }  ]] & := &  [[rho i]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

% Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
% syntax, including renaming, substitution, and operations over
% the typing context or substitution.

The \lang language is expressive enough to support large
eliminations, the ability to compute a type using a term as input. For
example, the function $[[\ Bool if var 0 Bool Bool -> Bool]]$ returns
either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
is $[[true]]$ or $[[false]]$.

We adapt from \citet{autosubst2} the notations for simultaneous
renaming, substitution, and other auxiliary definitions used in
our paper, summarized in Figure~\ref{fig:auxdef}. We need those
definitions to specify the reduction relation and the typing relation
for \lang{}. The metavariable $[[rho]]$ represents substitutions,
mappings from de Bruijn indices to terms.
One example of a substitution is
the function $\ottkw{id}_{tm}$, which takes a de Bruijn index directly
injects it as a term variable. The $\uparrow^{[[n]]}$ operator is
more general and adds $[[n]]$ to the index before injecting it as a
term. The $[[(rho .: a)]]$ operation allows us to extend a substitution
$[[rho]]$ with an extra element $[[a]]$; the extended substitution maps the variable
$[[var 0]]$ to $[[a]]$ and variables $[[var Suc i]]$ to $[[rho i]]$.

The substitution operator, which takes the form $[[a {rho}]]$,
traverses the syntax of $[[a]]$ and replaces each variable $[[var i]]$
with the term $[[rho i]]$. When traversing under binders (e.g. in the
$[[(\ a) { rho }]]$ case), the $\Uparrow$ operator prevents $[[var
0]]$, the bound variable, from being replaced by
$[[rho]]$. From the definition of $[[up rho]]$, we see that it keeps
$[[var 0]]$ unchanged during substitution and replaces each $[[var
Suc i]]$ with $[[rho i {up 1}]]$. The extra shifting with $[[up 1]]$
is required so the variables in terms from the codomain of $[[rho]]$ skip over
the newly introduced binder to refer to the correct binding
location. The substitution operator is referred to as simultaneous
substitution as it substitutes all variables at once. It is possible
to recover single substitution by composing the extension operator and
the identity substitution: $[[a { b }]] := [[a { idtm .: b }]]$.
